---
title: "Blog Post 3 - Mr. Trash Wheel Data"
author: "Jeffrey Merselis"
date: "2024-3-8"
categories: [code, analysis]
toc: true
---

# Introduction

* what data set are you working with?
* what variables are you most interested in? How many observations are in your data set?
* what is your question of interest?
* where did you get the data from? (If possible, give a link to the data source).
* what other pieces of context are important to explain to the reader?

```{r}
#| warning: false
library(tidyverse)
```

```{r}
#| warning: false
trashwheel <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-03-05/trashwheel.csv')
```

# Primary Visualizations

* choose an appropriate visualization, eliminate clutter, and focus attention as appropriate to answer your question of interest.
* provide a written interpretation of each visualization you have, explaining what a reader should “take away” from each visualization.

# Conclusion and Wrap-Up

* are there any flaws with your approach to analyzing the data you’ve selected?
* are there future directions you would go if you had additional data or more time?

# Connection to Class Ideas

* At the end of your write-up, you should connect your visualization(s) to ideas from class. In what ways would you argue your visualization(s) are effective ways of communicating information?